`plot.dads`

### 2023-04-05 inst/gitbook/04_make.events.Rmd

One other major feature of the `dads` package is to allow simulations to run with specified events to occur during the simulation.
These are typically events that can drastically change the course of the simulation.
For example, you might want simulate some mass extinction at some specific point in time.
However, these can also be more subtle, like the internal change of parameter values when reaching a specific trait value.
These are all handled by `dads` with the `events` object that you can create using `make.events` and has the same overall logic as `make.traits` and `make.modifiers`.

`events` require three main arguments:
 * the `target` which designates what the extinction should affect (e.g. the taxa, the speciation rate, etc...);
 * the `condition` which designates when to trigger the event;
 * the `modification` which designates what to exactly modify when the event is triggered.

@@

The target of the event is what the event is going to modify in the birth-death algorithm.
You can only have one target per event (along with one condition and one modification) but you can create events that contain multiple events (i.e. multiple triplets of target/condition/modification).
The targets that are currently available are:
 * `"taxa"` to modify anything linked to the [`lineage` list](#allowarguments). E.g. making half of the living taxa go extinct.
 * `"bd.params"` to modify anything linked to the [`bd.params` object](#makebdparams). For example you might want to change the distribution of one of the parameter after some `conditions`. This is typically done by updating the object using the argument `update` from `make.bd.params`.
 * `"traits"` to modify anything linked to the [`traits` object](#maketraits). For example you might want to change the trait process after some `conditions`. This is typically done by updating the object using the argument `update` from `make.traits`.
 * `"modifiers"` to modify anything linked to the [`modifiers` object](#makebdmodifiers). For example you might want to change the speciation rule after some `conditions`. This is typically done by updating the object using the argument `update` from `make.modifiers`.
 * `"founding"` this target is a bit more special and allows you run a nested `dads` object in the simulation. It is cover in a [specific section below](#founding).

We will see some example of these targets by illustrating the `conditions` and `modifications` below.

@@

All `condition` functions in `dads` take at least to arguments: `x` for the variable of interest (e.g. time, number of taxa, trait value, etc...) and `condition`, the relational operator to evaluate. A relational operator is the proper (fancy) term designating all the comparisons you're regularly using in `R` like `==` (is equal?) `<` (is smaller?) `>=` (is bigger or equal?). You can get the full list in the base manual (using `?Comparison` or any of the relational operator in a function form).

> Note on the function form of a operator in `R`: if you weren't aware, functions in `R` are _usually_ described in the format `function(arguments)`, however many other functions you are commonly using are written as operators, for example (`1 + 1`). These are still functions though! And you can always use them in the classic function format by quoting them in back ticks: the following does the same as `(1+1)`: `` `+`(1, 1) ``. You can use that for parsing your relational operators as functions in `condition`. If you want you condition to be equal, use the syntax ``condition = `==` ``. More details on how functions _really_ work in the [Advanced `R` book](https://adv-r.hadley.nz/functions.html)!

 * `time.condition` will trigger the `event` once a certain time is reached in the simulations and is the simplest/more basic condition with no arguments other than the time required. For example ``time.condition(4, condition = `==`)`` will trigger the `event` once the simulations reached 4 units of time. Easy.
 * `taxa.condition` will trigger the `event` once a certain number of taxa is reached. This can be considered including or excluding fossil species. For example ``taxa.condition(42, condition = `>=`, living = TRUE)`` will trigger the `event` if there is at least 42 living species.
 * `trait.condition` will trigger the `event` once a certain trait value is reached. This function allows to say which trait(s) to target (by default, the first one using `trait = 1`), what value of the trait to target (by default `what = max`) and whether to use and absolute trait value or not (`absolute = TRUE`). For example ``trait.condition(1/3, condition = `>`, trait = 1, what = sd)`` will trigger the condition after the standard deviation of the first trait reaches 1/3.

@@

After defining the `event`'s `target` and `condition`, you'll need to also specify what it should modify.
These are basically functions that should modify a specific aspect of the target
Depending on the target you can modify the following:

@@

The most straightforward example if for `modifications` on `bd.params`, `traits` or `modifiers` objects since they use the same syntax as for their generic `make.X` function.
For example, for `make.traits`, you can update a `trait` using the `update` argument as follows:

@@

So basically the functions `update.X` apply the update to the object `X` when the event happens.
The following `event` updates the `bd.params` object by setting the extinction parameter to 1/3 when reaching a 10 species:

```{r, eval = FALSE}
make.events(target       = "bd.params",
            condition    = taxa.condition(10, condition = `>=`),
            modification = update.bd.params(extinction = 1/3))

```

> Note that by default events are triggered only once in the whole simulation, so although the example above states that the condition is reaching at least 10 taxa, it will not trigger every times it reaches more than 10 taxa. You can change the number of times the events can be triggered using the argument `replications` (by default it's set to `replications = 0` for triggering it only once).

@@

Here are some example illustrating how to generate events.
For a simple example, we can create a extinction event that will remove 80% of species after reaching time 4:

@@

## Run the simulations
set.seed(1)
no_change <- dads(bd.params = bd.params,
                  stop.rule = stop.rule,
                  traits    = traits)
set.seed(1)
process_change <- dads(bd.params = bd.params,
                       stop.rule = stop.rule,
                       traits    = traits,
                       events    = change_process)
## Plot the results
par(mfrow = c(1,2))
plot(no_change, ylim = c(-7, 7))
plot(process_change, ylim = c(-7, 7))
```

@@

Founding events are a specific `target` for `events` that allows you to... simulate a birth-death process within the current one!
It basically allows you to simulate a specific `dads` process i.e. using the `dads` function with it's own `traits`, `modifiers`, `bd.params` and `events` (and yes, that's including `events` that have their own `founding` event)...
This is basically the ultimate nested boss of modularity!

The founding event will basically run an internal `dads` process (i.e. simulating a tree and, optionally, some data) resulting in a founding sub-tree.
It will then branch this sub-tree to the rest of the simulation that continued normally in the mean time.
You can specify the founding event using the inbuilt `founding.event` modification event.
This function intakes the exact same arguments as `dads` to simulate the sub-tree with its own parameters.
Additionally, we will use the `additional.args` argument from `make.events` to specify a prefix for the founding tree tips (to make them easier to distinguish).

@@

Note that the nestedness here is potentially endless, for example, you can parse an `event` argument to the `founding.event` that will generate another founding event, etc...

>>>>

One other major feature of the `dads` package is that it allows simulations to run with specified events that occur during the simulation.
These are typically events that can drastically change the course of the simulation.
For example, you might want simulate a mass extinction at some specific point in time.
However, these can also be more subtle, like the internal change of parameter values when reaching a specific trait value.
These are all handled by `dads` with the `events` object that you can create using `make.events` and has the same overall logic as `make.traits` and `make.modifiers`.

`events` require three main arguments:
 * the `target` which designates what the extinction should affect (e.g. the taxa, the speciation rate, etc...);
 * the `condition` which designates when to trigger the event;
 * the `modification` which designates what to modify when the event is triggered.

@@

The target of the event is what the event is going to modify in the birth-death algorithm.
You can only have one target per event (along with one condition and one modification) but you can create events that contain multiple events (i.e. multiple triplets of target/condition/modification).
The targets that are currently available are:
 * `"taxa"` to modify anything linked to the [`lineage` list](#allowarguments). e.g. making half of the living taxa go extinct.
 * `"bd.params"` to modify anything linked to the [`bd.params` object](#makebdparams). For example you might want to change the distribution of one of the parameter after some `conditions`. This is typically done by updating the object using the argument `update` from `make.bd.params`.
 * `"traits"` to modify anything linked to the [`traits` object](#maketraits). For example you might want to change the trait process after some `conditions`. This is typically done by updating the object using the argument `update` from `make.traits`.
 * `"modifiers"` to modify anything linked to the [`modifiers` object](#makebdmodifiers). For example you might want to change the speciation rule after some `conditions`. This is typically done by updating the object using the argument `update` from `make.modifiers`.
 * `"founding"` this target is a bit more specialised and allows you run a nested `dads` object in the simulation. It is covered in a [specific section below](#founding).

We will see some examples of these targets in the `conditions` and `modifications` described below.

@@

All `condition` functions in `dads` take at least two arguments: `x` for the variable of interest (e.g. time, number of taxa, trait value, etc.) and `condition`, the relational operator to evaluate. A relational operator is the proper (fancy) term designating all the comparisons you're regularly using in `R` like `==` (is equal?) `<` (is smaller?) `>=` (is bigger or equal?). You can get the full list in the R base manual (using `?Comparison` or any of the relational operator in a function form). For more details on how functions _really_ work in R see the [Advanced `R` book](https://adv-r.hadley.nz/functions.html)!

 * `time.condition` will trigger the `event` once a certain time is reached in the simulations and is the simplest/most basic condition with no arguments other than the time required. For example ``time.condition(4, condition = `==`)`` will trigger the `event` once the simulations reaches 4 time units. Easy.
 * `taxa.condition` will trigger the `event` once a certain number of taxa are reached. This can be considered including or excluding fossil species. For example ``taxa.condition(42, condition = `>=`, living = TRUE)`` will trigger the `event` if there are at least 42 living taxa.
 * `trait.condition` will trigger the `event` once a certain trait value is reached. This function allows to say which trait(s) to target (by default, the first one using `trait = 1`), what value of the trait to target (by default `what = max`) and whether to use and absolute trait value or not (`absolute = TRUE`). For example ``trait.condition(1/3, condition = `>`, trait = 1, what = sd)`` will trigger the condition after the standard deviation of the first trait reaches 1/3.

@@

After defining the `event`, `target` and `condition`, you also need to specify what it should modify.
These are functions that should modify a specific aspect of the target.
Depending on the target you can modify the following:

@@

The most straightforward example is for `modifications` on `bd.params`, `traits` or `modifiers` objects because they use the same syntax as for their generic `make.X` function.
For example, for `make.traits`, you can update a `trait` using the `update` argument as follows:

@@

So the function `update.X` applies the update to the object `X` when the event happens.
The following `event` updates the `bd.params` object by setting the extinction parameter to 1/3 when reaching a 10 species:

```{r, eval = FALSE}
make.events(target = "bd.params",
            condition = taxa.condition(10, condition = `>=`),
            modification = update.bd.params(extinction = 1/3))
```

> Note that by default events are triggered only once across the whole simulation, so although the example above states that the condition is reaching at least 10 taxa, it will not trigger every time it reaches more than 10 taxa, only the first time. You can change the number of times the events can be triggered using the argument `replications` (by default it's set to `replications = 0` for triggering the event only once).

@@

Here are some examples illustrating how to generate events.
For a simple example, we can create a extinction event that will remove 80% of species after reaching time 4:

@@

## Run the simulations
set.seed(1)
no_change <- dads(bd.params = bd.params,
                  stop.rule = stop.rule,
                  traits = traits)
set.seed(1)
process_change <- dads(bd.params = bd.params,
                       stop.rule = stop.rule,
                       traits = traits,
                       events = change_process)
## Plot the results
par(mfrow = c(1,2))
plot(no_change, ylim = c(-7, 7))
plot(process_change, ylim = c(-7, 7))
par(mfrow = c(1,1))
```

@@

Founding events are a specific `target` for `events` that allows you to simulate a birth-death process within the current one!
It allows you to simulate a specific `dads` process using the `dads` function with its own `traits`, `modifiers`, `bd.params` and `events` (and yes, that's including `events` that have their own `founding` event).
This is basically the ultimate nested boss of modularity!

The founding event will run an internal `dads` process (i.e. simulating a tree and, optionally, some data) resulting in a founding sub-tree.
It will then branch this sub-tree onto the rest of the simulation that continued normally in the mean time.
You can specify the founding event using the inbuilt `founding.event` modification event.
This function takes the exact same arguments as `dads` to simulate the sub-tree with its own parameters.
Additionally, we will use the `additional.args` argument from `make.events` to specify a prefix for the founding tree tips (to make them easier to distinguish).

@@

Note that the nestedness here is potentially endless, for example, you can pass an `event` argument to the `founding.event` that will generate another founding event, etc.

### 2023-04-05 inst/gitbook/06_plotting_dads_objects.Rmd
Update 06_plotting_dads_objects.Rmd

@@

# Plotting traits

@@

```{r, fig.height=8, fig.width=8}
## Making a list of three traits
list_of_traits <- make.traits(process = c(no.process, BM.process, OU.process), trait.names = c("No process (normal)", "Brownian motion", "Ornstein-Uhlenbeck"))
## Plotting each trait separately
par(mfrow = c(2, 2))
plot(list_of_traits, trait = 1)
## Using different colours options
plot(list_of_traits, trait = 2, col = c("red", "purple", "pink"))
## Not using the default plot name
plot(list_of_traits, trait = 3, main = "OU process",
     cent.tend = sd, quantiles = c(10, 30))
```

You can also control the number of replicates in the simulation by using the `simulations` option (the default is 50).
Bigger numbers leads to more time but smoother looking plots while smaller ones are more stochastic:

```{r, fig.width=6, fig.height=12}
par(mfrow = c(2,1))
plot(list_of_traits, trait = 2, simulations = 10, main = "10 BM simulations")
plot(list_of_traits, trait = 2, simulations = 1000, main = "1k BM simulations")
```

# Plotting dads results

If `dads` is used to plot only a tree (and outputs a `"phylo"` object), you can use the function `plot.phylo` from the `ape` package to plot your tree.
You'll find all the tree plotting option in the `?plot.phylo` manual page.

@@

By default, elements are coloured as follows: nodes and tips are points and coloured in blue if they are tips (light blue if they are fossils) and nodes are in orange.
Branches linking them are grey lines.
You can of course change this colour palette to something of you preference by calling the normal arguments that can be passed to `points` or `lines`. For example `pch` for the point type or `lty` for the line type.
Furthermore, time is plotted conventionally from left to right (left is towards the past, right is towards the present) but you can change that by specifying `xlim`.

```{r, fig.height = 6, fig.width = 6}
## Playing with some more options
plot(my_tree, main = "A tree and traits",
     ## Changing nodes colours and type
     col = c(tips = "pink", nodes = "purple"), pch = 21,
     ## Changing edges colour and type
     lwd = 2, lty = 3, edges = "yellow",
     ## Changing the x axis orientation and label
     xlim = c(0, 2.5), xlab = "Time into the past")
```

> Note that you can modify colours using the `col` argument by providing a clear indication of what you want to colour (e.g. `col = c(tips = "blue", fossils = "orange")` will apply the colours to the unambiguously named elements) or by providing a function that will scale the colours with time (e.g. `col = grDevices::heat.colors` will use the specified function to change the elements colours with time).

And you can add a default legend by using `legend = TRUE` (if you don't want to add the default legend you can add it after your plot using `legend(...)`).

@@

## 3D version!

@@

```{r, fig.height = 6, fig.width = 6}
## Specifying a 3D trait process
my_3D_trait <- make.traits(n = 3, )
## Simulating a birth death tree with that 3D trait
my_data <- dads(bd.params = list(extinction = 0.2),
                stop.rule = list(max.living = 50),
                traits    = my_3D_trait)
```

You can toggle which trait to plot using the `trait` option.
Either by providing a single value to plot that specific trait against time or by providing two traits.

@@

If two traits are provided, the 3rd dimensions is going to be time by default:

>>>>

# Plots

@@

## Plotting traits

@@

```{r, fig.height=8, fig.width=8}
## Making a list of three traits
list_of_traits <- make.traits(process = c(no.process, BM.process, OU.process), trait.names = c("No process (normal)", "Brownian motion", "Ornstein-Uhlenbeck"))
## Plotting each trait separately
par(mfrow = c(2, 2))
plot(list_of_traits, trait = 1)
## Using different colours options
plot(list_of_traits, trait = 2, col = c("red", "purple", "pink"))
## Not using the default plot name
plot(list_of_traits, trait = 3, main = "OU process",
     cent.tend = sd, quantiles = c(10, 30))
par(mfrow = c(1, 1))
```

You can also control the number of replicates in the simulation by using the `simulations` option (the default is 50).
Bigger numbers lead to more time but smoother looking plots while smaller ones are more stochastic:

```{r, fig.width=6, fig.height=12}
par(mfrow = c(2,1))
plot(list_of_traits, trait = 2, simulations = 10, main = "10 BM simulations")
plot(list_of_traits, trait = 2, simulations = 1000, main = "1k BM simulations")
par(mfrow = c(1,1))
```

## Plotting `dads` results

If `dads` is used to plot only a tree (and outputs a `"phylo"` object), you can use the function `plot.phylo` from the `ape` package to plot your tree.
You'll find all the tree plotting options in the `?plot.phylo` manual page.

@@

By default, elements are coloured as follows: nodes and tips are points and coloured in blue if they are tips (light blue if they are fossils) and nodes are in orange.
Branches linking them are grey lines.
You can of course change this colour palette to whatever you prefer by calling the standard base R arguments that can be passed to `points` or `lines`. For example `pch` for the point type or `lty` for the line type.
Furthermore, time is plotted conventionally from left to right (left is towards the past, right is towards the present) but you can change that by specifying `xlim`.

```{r, fig.height = 6, fig.width = 6}
## Playing with some more options
plot(my_tree, main = "A tree and traits",
     ## Changing nodes colours and type
     col = c(tips = "pink", nodes = "purple"), pch = 21,
     ## Changing edge colour and type
     lwd = 2, lty = 3, edges = "yellow",
     ## Changing the x axis orientation and label
     xlim = c(0, 2.5), xlab = "Time into the past")
```

> Note that in `plot.dads` you can modify colours using the `col` argument by providing a clear indication of what you want to colour (e.g. `col = c(tips = "blue", fossils = "orange")` will apply the colours to the unambiguously named elements) or by providing a function that will scale the colours with time (e.g. `col = grDevices::heat.colors` will use the specified function to change the  colours of the elements through time).

You can add a default legend by using `legend = TRUE` (if you don't want to add the default legend you can add it after your plot using `legend(...)`).

@@

### Adding dimensions! Not necessary, but fun!

@@

```{r, fig.height = 6, fig.width = 6}
## Specifying a 3D trait process
my_3D_trait <- make.traits(n = 3, )
## Simulating a birth-death tree with that 3D trait
my_data <- dads(bd.params = list(extinction = 0.2),
                stop.rule = list(max.living = 50),
                traits    = my_3D_trait)
```

You can toggle which trait to plot using the `trait` option, either by providing a single value to plot that specific trait against time or by providing two traits.

@@

If two traits are provided, the 3rd dimension is going to be time by default:
